---
title: "Time to Restore - Summary of Available Data"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Erin Zylstra"
execute: 
  echo: false
  error: false
  warning: false
  message: false
format:
  html:
    embed-resources: false
    fig-cap-location: top
    page-layout: full
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r packages-settings}

require(rnpn)
require(lubridate)
require(stringr)
require(dplyr)
require(tidyr)
# require(sf)
require(terra)
require(ggplot2)
require(leaflet)
# require(map)
# require(maptiles)
require(tidyterra)
require(jsonlite)
require(knitr)
require(kableExtra)

# Logical indicating whether to re-download data
update_data <- FALSE
```

### Identifying priority species

Used list of priority species based on: TimeToRestore_AllPrioritySpecies_2024 [google sheet](https://docs.google.com/spreadsheets/d/1NnuTGT7OZuA9WGD0cg29d3INNV0n_x9C/edit?gid=794771851#gid=794771851) (extracted on 2 April 2025) and made the following changes:

- Corrected the spelling of one genus (*Vernonia*) and one species (*cespitosa*)

- Updated scientific and common names based on ITIS for mistflower (*Conoclinium*), basket-flower (*Centaurea*), and *Lantana*; will use the common name listed in ITIS/NPN data base for these species imoving forward.

- Using the common name listed in NPN database for *Sambucus nigra* and *Passiflora incarnata*, which were each listed twice in the googlesheet under two different common names.

```{r priority-spp}
#| code-fold: true

# Load list of priority species
spp_list <- read.csv("data/ttr-priorityspecies-20250402.csv", 
                     na.strings = c(NA, ""))
spp_list <- spp_list %>%
  mutate(across(c(common_name, scientific_name), str_trim)) %>%
  # Edit spelling of one genus (Vernonia) and one species (cespitosa)
  # Update species based on ITIS for mistflower (Conoclinium), 
  # basket-flower (Centaurea), and Lantana
  mutate(scientific_name = case_when(
    scientific_name == "Oenothera caespitosa" ~ "Oenothera cespitosa",
    scientific_name == "Veronia gigantea" ~ "Vernonia gigantea",
    scientific_name == "Conoclinium greggii" ~ "Conoclinium dissectum",
    scientific_name == "Centaurea americana" ~ "Plectocephalus americanus",
    scientific_name == "Lantana urticoides" ~ "Lantana horrida", 
    .default = scientific_name
  )) %>%
  rename(ttr_common_name = common_name)

# Load information about NN species
nn_spp <- npn_species() %>% data.frame()
nn_spp <- nn_spp %>%
  filter(kingdom == "Plantae") %>%
  select(species_id, common_name, genus, species, functional_type) %>%
  mutate(scientific_name = paste(genus, species))

# Find NN info based on scientific name of priority species (inconsistent 
# capitalization in priority list and a duplicate in NN database [Canada goldenrod])
spp_list <- spp_list %>%
  left_join(nn_spp, by = "scientific_name")
# Check that all priority species have a match in NN database
  # filter(spp_list, is.na(species_id))
# Are there any duplicates? Yes
  # count(spp_list, species_id) %>% filter(n > 1)
  # filter(spp_list, species_id == 90)
  # In priority list, Sambucus nigra listed as both black and common elderberry
  # filter(spp_list, species_id == 182)
  # In priority list, Passiflora incarnata listed as both purple passionflower and Maypop

# Remove entries for TTR priority species whose common name doesn't match NPN
spp_list <- spp_list %>%
  filter(!ttr_common_name %in% c("Maypop", "Common elderberry"))
```

This left us with `r nrow(spp_list)` priority species.

```{r load-states}

# File location
states <- vect("data/states/cb_2017_us_state_500k.shp")

```

```{r download}
#| code-fold: true
#| output: false

# First, check that all species have these 4 flower/fruit phenophases
phenophases_byspp <- npn_phenophases_by_species(
  species_ids = c(spp_list$species_id),
  date = "2026-04-01"
) %>% data.frame()
phenophases_byspp %>%
  group_by(species_id, species_name) %>%
  summarize(p500 = ifelse(500 %in% phenophase_id, 1, 0),
            p501 = ifelse(501 %in% phenophase_id, 1, 0),
            p516 = ifelse(516 %in% phenophase_id, 1, 0),
            p390 = ifelse(390 %in% phenophase_id, 1, 0),
            .groups = "drop") %>%
  rowwise() %>%
  filter(sum(c_across(p500:p390)) < 4)
# All species use these 4 phenophases

# Will skip any leaves phenophases for now, though we may want to add them back
# in, if only for Asclepias spp.

# Download and format (or load existing) NPN data for priority plant species --#
  
# We will focus on observations from 4 states (LA, NM, OK, TX)
# We'll also restrict observations to those collected in 2013 to present, since
# different protocols and intensity categories will used for some species before
# then.

phenophases <- c(500, 501, 516, 390)
states4 <- c("LA", "NM", "OK", "TX")
years <- 2013:2026
  # Note: we could be missing observations in the four states if we use
  # the states argument in the download function because sometimes the state 
  # field is missing or incorrect. Best to download all records for species and
  # fixing/imputing state, and then filtering by state.

data_filename <- "data/status-intensity-May2026.csv"

if (!file.exists(data_filename) | update_data == TRUE) {
  
  # Download flowering, fruiting data including all years and not just years 
  # when TTR project existed (to take advantage of all observations)
  status_dl <- npn_download_status_data(
    request_source = "erinz",
    years = years,
    species_ids = spp_list$species_id,
    phenophase_ids= phenophases,
    additional_fields = c("observedby_person_id",
                          "partner_group",
                          "site_name"))
  status_dl <- data.frame(status_dl)
  
  # Format data
  status <- status_dl %>%
    mutate(obsdate = ymd(observation_date),
           yr = year(obsdate),
           php = case_when(
             phenophase_id == 500 ~ "flower",
             phenophase_id == 501 ~ "open flower",
             phenophase_id == 516 ~ "fruit",
             phenophase_id == 390 ~ "ripe fruit")) %>%
    select(-c(update_datetime, elevation_in_meters, genus, species, kingdom,
              phenophase_description, abundance_value, observation_date)) %>%
    rename(person_id = observedby_person_id,
           lat = latitude,
           lon = longitude)
  
  # Some observations missing state ID. Will use a shapefile to get an assigned
  # state for each site and use that moving forward
  state_fill <- status %>%
    select(site_id, lon, lat, state) %>%
    distinct()
  state_fillv <- vect(state_fill, 
                      geom = c("lon", "lat"), 
                      crs = "epsg:4326")
  state_new <- terra::extract(states, state_fillv)
  state_fill <- cbind(state_fill, state_new = state_new$STUSPS)
  # check:
  # count(state_fill, state, state_new) %>%
  #   mutate(same = ifelse(state == state_new, 1, 0)) %>%
  #   arrange(same)
  
  #### Need to decide whether we want to retain observations of these species
  #### in just the 4-state area or whether we want to expand that a little bit. 
  #### One option would be to widen the region in each direction, so include:
  #### AZ, UT, CO, KS, MO, AR, MS, AL (in addition to 4 focal states). 
  region <- c(states4, "AZ", "UT", "CO", "KS", "MO", "AR", "MS", "AL")
  
  # Attach new state labels and exclude observations that aren't in the 4 states
  status <- status %>%
    left_join(select(state_fill, site_id, state_new), by = "site_id") %>%
    select(-state) %>%
    rename(state = state_new) %>%
    filter(!is.na(state) & state %in% region)
  
  # Write to file
  write.csv(status, data_filename, row.names = FALSE)
  # Remove objects
  rm(status, status_dl)
}

df <- read.csv(data_filename) %>%
  mutate(obsdate = ymd(obsdate))
```

```{r data-cleanup}

# Remove any observations with phenophase_status = -1 (even if intensity value
# was reported). This is <1% of observations.
df <- df %>%
  filter(phenophase_status != -1)

# Convert intensity value ranges to midpoints
df <- df %>%
  mutate(midpoint = case_when(
    intensity_value == "Less than 3" ~ 2, 
    intensity_value == "3 to 10" ~ ceiling(mean(c(3, 10))), 
    intensity_value == "11 to 100" ~ ceiling(mean(c(11, 100))), 
    intensity_value == "101 to 1,000" ~ ceiling(mean(c(101, 1000))), 
    intensity_value == "1,001 to 10,000" ~ ceiling(mean(c(1001, 10000))), 
    intensity_value == "More than 1,000" ~ 1001, 
    intensity_value == "More than 10,000" ~ 10001, 
    intensity_value == "Less than 5%" ~ 0.03, 
    intensity_value == "5-24%" ~ ceiling(mean(c(5, 24))) / 100, 
    intensity_value == "25-49%" ~ ceiling(mean(c(25, 49))) / 100, 
    intensity_value == "50-74%" ~ ceiling(mean(c(50, 74))) / 100, 
    intensity_value == "75-94%" ~ ceiling(mean(c(75, 94))) / 100, 
    intensity_value == "95% or more" ~ 0.95,
    .default = NA)
  )
# Check that status equals 1 if there's an intensity value
# count(df, phenophase_status, midpoint)

# Clean up columns, column names
df <- df %>%
  select(-intensity_category_id) %>%
  rename(person = person_id,
         lpp = partner_group,
         site = site_id,
         spp_id = species_id,
         spp = common_name, 
         id = individual_id, 
         php_id = phenophase_id, 
         doy = day_of_year,
         status = phenophase_status)

# Just going to look at flowering data for now
df <- df %>%
  filter(str_detect(php, "flower"))
```

### Summarizing where flowering data have been collected

Mapping out some potential regions to subset observations in Texas (size indicates number of plants monitored at that site, 2013-2026)

```{r map-potential-regions}

locs <- df %>%
  # Limit to one observation per plant per day
  distinct(site, lat, lon, spp_id, spp, id, doy, obsdate, yr, state) %>%
  group_by(state, site, lat, lon) %>%
  summarize(n_spp = n_distinct(spp),
            n_plants = n_distinct(id),
            n_obs = n(),
            n_obs_2325 = sum(yr %in% 2023:2025),
            .groups = "drop") %>%
  data.frame()

# Site = no. plants  
map_nplants <- locs %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    weight = 2,
    radius = ~n_plants)

# Size = no. observations in 2023-2025 (handfull of dominant sites!)
map_nobs <- locs %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    weight = 2,
    radius = ~n_obs_2325/50)

# What should regions be?
# States: NM, OK, 
# All of LA? (only a few obs in northern half). Lots of observations along
  # the coast in MS and AL that might be helpful.
# TX: Houston area: 29.736183, -95.332086
# TX: Austin - San Antonio area (include a couple big sites west of Waco?)
  # San Antonio: 29.432180, -98.488961
  # Austin: 30.271591, -97.735506
  # Between (San Marcos): 29.892405, -98.039982
# TX: Ft Worth/Dallas area: 32.771473, -96.996293
tx_cities <- data.frame(
  city = c("Houston", "Austin", "San Antonio", "San Marcos", "Dallas"),
  lat = c(29.761647, 30.271591, 29.447646, 29.892405, 32.771473),
  lon = c(-95.391512, -97.735506, -98.605101, -98.039982, -96.996293)
)

# Testing out some 100-km radius areas in TX
locs %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    data = filter(locs, state %in% c("LA", "TX", "NM", "OK")),
    lng = ~lon,
    lat = ~lat,
    weight = 2,
    radius = ~n_plants) %>%
  addCircleMarkers(
    data = filter(locs, !state %in% c("LA", "TX", "NM", "OK")),
    lng = ~lon,
    lat = ~lat,
    weight = 2,
    color = "gray",
    radius = ~n_plants) %>%
  addCircles(
    data = filter(tx_cities, city != "San Marcos"),
    lat = ~lat,
    lng = ~lon,
    weight = 2,
    radius = 100000,
    color = "red", 
    fill = NA)

map_nobs_regions <- locs %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    weight = 2,
    color = "purple",
    radius = ~n_obs_2325/50) %>%
  addCircles(
    data = filter(tx_cities, city != "San Marcos"),
    lat = ~lat,
    lng = ~lon,
    weight = 2,
    radius = 100000,
    color = "red")
```

\

Decided on 3 regions
```{r finalize-regions}

# Create buffers and merge the Austin/San Antonio buffers
tx_citiesv <- tx_cities %>% 
  filter(city != "San Marcos") %>%
  vect(geom = c("lon", "lat"), "epsg:4326")
tx_regions <- buffer(tx_citiesv, width = 100000)
tx_regions <- aggregate(tx_regions) %>% disagg()

# Find which point each dissolved polygon overlaps with
overlap_matrix <- relate(tx_regions, tx_citiesv, relation = "intersects")
# For each dissolved polygon, get the index of the first matching point
first_match <- apply(overlap_matrix, 1, function(x) which(x)[1])
# Subset points by those indices to get attributes
attrs <- as.data.frame(tx_citiesv)[first_match, ]
# Attach attributes to dissolved polygons
values(tx_regions) <- attrs
# Rename attribute
names(tx_regions) <- "region"
# Check:
# ggplot() +
#   geom_spatvector(data = tx_regions, aes(fill = region))

# Create a region assignment for each location
locsv <- vect(locs, geom = c("lon", "lat"), "epsg:4326") 
regions <- terra::extract(tx_regions, locsv)
locs$region <- regions$region

locs <- locs %>%
  mutate(state4 = ifelse(state %in% c("LA", "TX", "NM", "OK"), 1, 0),
         state5 = ifelse(state4 == 0, "Outside", state))

# Map check:
ggplot() +
  geom_spatvector(data = subset(states, states$STUSPS %in% c("LA", "TX", "NM", "OK")),
                  fill = "white") +
  geom_spatvector(data = tx_regions, color = "red", fill = NA) +
  geom_point(data = filter(locs, state %in% c("LA", "TX", "NM", "OK")),
                           aes(x = lon, y = lat, color = region)) +
  labs(color = "Region") +
  theme(axis.title = element_blank())

# Summarize information for each state:
state_summary <- locs %>% 
  group_by(state5) %>%
  summarize(sites = n(),
            plants = sum(n_plants),
            obs = sum(n_obs_2325)) %>%
  data.frame()

# Summarize information for each region:
region_summary <- locs %>% 
  group_by(region) %>%
  summarize(sites = n(),
            plants = sum(n_plants),
            obs = sum(n_obs_2325)) %>%
  data.frame()
```

Flowering data by state (inside/outside 4-state area)

```{r state-summary}

state_summary %>%
  mutate(state5 = factor(state5, levels = c("LA", "NM", "OK", "TX", "Outside"))) %>%
  arrange(state5) %>%
  kable(col.names = c("Location", "Total no. sites", "Total no. plants",
                      "No. observations (2023-2025)"))

```

\

Flowering data in each Texas region (vs everywhere else)

```{r region-summary}

region_summary %>%
  kable(col.names = c("Region", "Total no. sites", "Total no. plants",
                      "No. observations (2023-2025)"))

```

```{r save}

# Attach region information to flowering data
df <- df %>%
  left_join(select(locs, site, region, state4, state5), by = "site")

# Write observation data and region boundaries to file
flowers_filename <- "data/status-intensity-flowers-May2026.csv"

if (!file.exists(flowers_filename) | update_data == TRUE) {
  write.csv(df, "data/status-intensity-flowers-May2026.csv", row.names = FALSE)
  writeVector(tx_regions, "data/tx-regions/tx-regions.shp")
}
```
